### 2010-10-11 Thesis/Chapter4.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}). Another illustration of this stability is shown \fig{chap4:fig-duck}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time. Image courtesy of \cite{Muller05}.}
\label{chap4:fig-shoes}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
\vspace{0.1cm}
\includegraphics[width=11cm]{chapter4/duck2.pdf}
\end{center}
\caption[Squeezing of a duck using shape matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique. Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations. Image courtesy of \cite{Muller05}}
\label{chap4:fig-duck}
\end{figure}

@@

	\subsection{3D Chainmail algorithm}
The \emph{3D Chainmail} was introduced by \cite{Gibson97} as a fast algorithm for deforming solid objects. Rather than carrying out complex computations on a low resolution geometry (as opposed to FEM techniques), the main idea is to take advantage of the original data resolution (of a CT scan for instance) but perform simple deformation calculation for each element. In this algorithm, each volume element is linked to its six nearest neighbours. If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links. It works in a very similar way to a chain and is illustrated by \fig{chap4:fig-chainmail}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=10cm]{chapter4/chainmail.pdf}
\end{center}
\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved. Image courtesy of \cite{Gibson97}.}
\label{chap4:fig-chainmail}
\end{figure}	
%
The chain mail algorithm is then followed by an elastic relaxation, which relaxes the shape of the deformed object approximated by the chain mail algorithm. This step consists of minimising an elastic energy measure within the object to insure that the calculated deformation is a minimum energy configuration.

@@

Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. \TODO{add PhD of Brian Lee... Mais je l'ai pas sa thèse... et tu m'as pas dit que tu étais en train de la reviewer ? Du coup c'est pas encore publique, c'est difficile de la citer, non ?}
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
\end{center}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

As we know, the assumption of linearity is usually only valid for very small deformations and strains. If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations. However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods \label{chap4:corotationalMethods}. A very complete description of the co-rotational approach is given by \cite{Felippa00}. As explained in section~\ref{chap2:strainMeasures}, an appropriate strain measure is a tensor which only measures the deformation applied to the object. This is the case of the Cauchy-Green or Green tensors for instance. Those strain metrics only measures deformation, they are invariant with respect to rigid-body transformations applied. However, geometrically linear analyses make use of the infinitesimal strain tensor, which is only an approximation of the Green tensor. Consequently, this linearised strain measure is not invariant with respect to rigid-body transformations. In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element. This error in the measure creates ghost forces which overdeforms the simulated object. The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components. In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure. Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear).

@@

The first geometrically non-linear finite element model running in real-time was reported by \cite{Debunne01}. They provided an adaptative method for animating viscoelastic deformable objects in a guaranteed frame rate. As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see \fig{chap4:fig-gummyBear}). They reported animating a few hundred points in real-time at a frequency higher than $ 300\,$Hz.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/gummyBear.pdf}
\end{center}
\caption[Simulation of a gummy bear]{The approach of \cite{Debunne01} makes use of a local refinement technique to ensure high physical fidelity while bounding the global computation load to guarantee real-time simulation with a geometrically non-linear and visco-elastic finite element model.} 
\label{chap4:fig-gummyBear}
\end{figure}

>>>>

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}). Another illustration of this stability is shown \fig{chap4:fig-duck}. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time. Image courtesy of \cite{Muller05}.}
\label{chap4:fig-shoes}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
\vspace{0.1cm}
\includegraphics[width=11cm]{chapter4/duck2.pdf}
\end{center}
\caption[Squeezing of a duck using shape matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique. Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations. Image courtesy of \cite{Muller05}}
\label{chap4:fig-duck}
\end{figure}

@@

	\subsection{3D Chainmail algorithm}
The \emph{3D Chainmail} was introduced by \cite{Gibson97} as a fast algorithm for deforming solid objects. Rather than carrying out complex computations on a low resolution geometry (as opposed to FEM techniques), the main idea is to take advantage of the original data resolution (of a CT scan for instance) but perform simple deformation calculation for each element. In this algorithm, each volume element is linked to its six nearest neighbours. If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links. It works in a very similar way to a chain and is illustrated by \fig{chap4:fig-chainmail}. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=10cm]{chapter4/chainmail.pdf}
\end{center}
\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved. Image courtesy of \cite{Gibson97}.}
\label{chap4:fig-chainmail}
\end{figure}	
%
The chain mail algorithm is then followed by an elastic relaxation, which relaxes the shape of the deformed object approximated by the chain mail algorithm. This step consists of minimising an elastic energy measure within the object to insure that the calculated deformation is a minimum energy configuration.

@@

Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. \TODO{add PhD of Brian Lee... Mais je l'ai pas sa thèse... et tu m'as pas dit que tu étais en train de la reviewer ? Du coup c'est pas encore publique, c'est difficile de la citer, non ?}
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
\end{center}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

As we know, the assumption of linearity is usually only valid for very small deformations and strains. If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations. However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods \label{chap4:corotationalMethods}. A very complete description of the co-rotational approach is given by \cite{Felippa00}. As explained in section~\ref{chap2:strainMeasures}, an appropriate strain measure is a tensor which only measures the deformation applied to the object. This is the case of the Cauchy-Green or Green tensors for instance. Those strain metrics only measures deformation, they are invariant with respect to rigid-body transformations applied. However, geometrically linear analyses make use of the infinitesimal strain tensor, which is only an approximation of the Green tensor. Consequently, this linearised strain measure is not invariant with respect to rigid-body transformations. In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element. This error in the measure creates ghost forces which overdeforms the simulated object (see \fig{chap4:fig-cylinder}). The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components. In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure. Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear). 
%
\begin{figure}[ht]
\centering 
\includegraphics[height=6cm]{chapter4/cylinder1.pdf}
\hspace{0.5cm}
\includegraphics[height=6cm]{chapter4/cylinder2.pdf}
\caption[Deformation of cylinders]{A geometrically linear finite element model (green) is compared to a co-rotational finite element model (yellow). While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not overdeform.}
\label{chap4:fig-cylinder}
\end{figure}

@@

The first geometrically non-linear finite element model running in real-time was reported by \cite{Debunne01}. They provided an adaptative method for animating viscoelastic deformable objects in a guaranteed frame rate. As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see \fig{chap4:fig-gummyBear}). They reported animating a few hundred points in real-time at a frequency higher than $ 300\,$Hz.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13cm]{chapter4/gummyBear.pdf}
\end{center}
\caption[Simulation of a gummy bear]{The approach of \cite{Debunne01} makes use of a local refinement technique to ensure high physical fidelity while bounding the global computation load to guarantee real-time simulation with a geometrically non-linear and visco-elastic finite element model.} 
\label{chap4:fig-gummyBear}
\end{figure}

### 2010-10-11 Thesis/Chapter6.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\begin{figure}
\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA. \textbf{Left:} a Behaviour Model controls the other representations via a series of mappings. \textbf{Right:} examples of representations with a liver model.}
\label{chap6:fig-SOFA}
\end{figure}

@@

% \TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}

@@

In \fig{chap6:fig-stressCurvesCube1} we demonstrate the effects of material anisotropy independent of viscoelastic effects by plotting the limiting instantaneous (hyperelastic) stress curves, produced by setting the relaxation parameter $\alpha_1 = 0$. \fig{chap6:fig-stressCurvesCube1}(a) shows curves for the anisotropic model, while \fig{chap6:fig-stressCurvesCube1}(b) shows curves for an equivalent isotropic model (i.e. with $\eta = 0$). Firstly we confirm that the FE solutions (data marks) match the analytical solutions (solid curves) very accurately. Secondly we observe that the anisotropic term significantly affects both the magnitude and shape of the stress curves. In particular the $S_{11}$ curve which is coincident with the $S_{33}$ curve in the isotropic model actually changes sign when anisotropy is introduced. The other curves exhibit a more pronounced stiffening for the anisotropic case. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter6/stressCurvesCube1.pdf}
\end{center}
\caption[Stress comparison with analytical solution for a pure shear of a cube]{Instantaneous stress curves for pure shear deformation of (a) the anisotropic model compared with those of (b) an isotropic model. Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
\label{chap6:fig-stressCurvesCube1}
\end{figure}

In \fig{chap6:fig-stressCurvesCube2} we demonstrate the effect of loading rate on the complete viscoelastic model by plotting the $S_{11}$ component for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$, plus the bounding instantaneous and equilibrium responses. The strain rate-dependence introduced by the model (and commonly observed in biological tissues) is clearly shown. Additionally the close match between the analytic and numerical solutions demonstrates the validity of the developed constitutive update scheme.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressCurvesCube2.pdf}
\end{center}
\caption[Stress comparison with analytical solution for a pure shear of a cube at various strain rate]{$S_{11}$ curves for pure shear deformation of the anisotropic viscoelastic model at varying strain rates. Curves for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$ are given (as labelled), along with the bounding instantaneous and equilibrium responses. The latter two are labelled $ \infty $ and $0^+$, respectively, indicating that they correspond to strain rates approaching these values. Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
\label{chap6:fig-stressCurvesCube2}
\end{figure}

@@

Figure \ref{chap6:fig-cubeAnisotropic}(a) shows the undeformed cube, while \fig{chap6:fig-cubeAnisotropic}(b) shows the deformed isotropic model. With no preferred direction the lateral expansion was uniform. In \fig{chap6:fig-cubeAnisotropic}(c) the increased $y$-direction stiffness of the first anisotropic model lead to a much reduced expansion in this direction (vertical in the image), and an accompanying increase in the orthogonal $z$-direction. For the second anisotropic model the direction defined by $y = z$ was stiffened, and \fig{chap6:fig-cubeAnisotropic}(d) shows the resulting reduced expansion along this axis and the increased orthogonal expansion. 
%
\begin{figure}[h]
\centering 
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic1.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic2.pdf}} \\
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic3.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic4.pdf}} \\
\caption[Deformation patterns of transversely isotropic models compared with that of an isotropic model]{Deformation patterns of transversely isotropic models compared with that of an isotropic model: (a) the undeformed cube, (b) the deformed isotropic model, (c) the deformed anisotropic model with $ \mathbf{a}_0 = \mathbf{a}_0^y $, (d) the deformed anisotropic model with $\mathbf{a}_0^{yz}$. Colour maps indicate relative magnitude of lateral displacement  $((u_y^2 + u_z^2)^{1/2})$.}
\label{chap6:fig-cubeAnisotropic}
\end{figure}

@@

The end face reaction force history for each model is shown in \fig{chap6:fig-stressDecay}. A distinct decay curve resulting from stress relaxation, and similar to that commonly noted in biological tissues \citep{Fung93}, was observed. The increased stiffness afforded by the anisotropic models (albeit in an orthogonal direction to the loading) result in greater reaction forces than the isotropic model.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressDecay.pdf}
\end{center}
\caption[FE computed reaction forces on the end faces of the cube models over time.]{FE computed reaction forces on the end faces of the cube models over time. Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$. Note that in this figure both markers and solid lines correspond to FE solutions.}
\label{chap6:fig-stressDecay}
\end{figure}

@@

		\subsubsection*{Performance within SOFA}
The efficiency and the side-effects of porting the algorithm into the flexible SOFA framework need to be measured. Therefore the performance has been assessed by comparing the computational time of the algorithm running within and outside SOFA. NVIDIA provides a tool to check the GPU implementation by evaluating many variables during the execution like for instance timings, counts of inconsistent reads and writes or GPU occupancy. We used this tool to carry out two measures:
\begin{enumerate}
\item GPU time only estimates the GPU computational time.
\item CPU time allows the evaluation of the execution time with the additional overhead due to the framework.
\end{enumerate}
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cube_and_curves1.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=8cm]{chapter6/cube_and_curves2.pdf}}
\caption [Computational timings within SOFA] {(a) FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load. (b) Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes. \TODO{change the cube if I can get a decent visual model in SOFA}}
\label{chap6:fig-cube_and_curves}
\end{figure}

>>>>

\begin{figure}[ht]
\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA. \textbf{Left:} a Behaviour Model controls the other representations via a series of mappings. \textbf{Right:} examples of representations with a liver model.}
\label{chap6:fig-SOFA}
\end{figure}

@@

% \{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}

@@

In \fig{chap6:fig-stressCurvesCube1} we demonstrate the effects of material anisotropy independent of viscoelastic effects by plotting the limiting instantaneous (hyperelastic) stress curves, produced by setting the relaxation parameter $\alpha_1 = 0$. \fig{chap6:fig-stressCurvesCube1}(a) shows curves for the anisotropic model, while \fig{chap6:fig-stressCurvesCube1}(b) shows curves for an equivalent isotropic model (i.e. with $\eta = 0$). Firstly we confirm that the FE solutions (data marks) match the analytical solutions (solid curves) very accurately. Secondly we observe that the anisotropic term significantly affects both the magnitude and shape of the stress curves. In particular the $S_{11}$ curve which is coincident with the $S_{33}$ curve in the isotropic model actually changes sign when anisotropy is introduced. The other curves exhibit a more pronounced stiffening for the anisotropic case. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=13.9cm]{chapter6/stressCurvesCube1.pdf}
\end{center}
\caption[Stress comparison with analytical solution for a pure shear of a cube]{Instantaneous stress curves for pure shear deformation of (a) the anisotropic model compared with those of (b) an isotropic model. Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
\label{chap6:fig-stressCurvesCube1}
\end{figure}

In \fig{chap6:fig-stressCurvesCube2} we demonstrate the effect of loading rate on the complete viscoelastic model by plotting the $S_{11}$ component for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$, plus the bounding instantaneous and equilibrium responses. The strain rate-dependence introduced by the model (and commonly observed in biological tissues) is clearly shown. Additionally the close match between the analytic and numerical solutions demonstrates the validity of the developed constitutive update scheme.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressCurvesCube2.pdf}
\end{center}
\caption[Stress comparison with analytical solution for a pure shear of a cube at various strain rate]{$S_{11}$ curves for pure shear deformation of the anisotropic viscoelastic model at varying strain rates. Curves for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$ are given (as labelled), along with the bounding instantaneous and equilibrium responses. The latter two are labelled $ \infty $ and $0^+$, respectively, indicating that they correspond to strain rates approaching these values. Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
\label{chap6:fig-stressCurvesCube2}
\end{figure}

@@

Figure \ref{chap6:fig-cubeAnisotropic}(a) shows the undeformed cube, while \fig{chap6:fig-cubeAnisotropic}(b) shows the deformed isotropic model. With no preferred direction the lateral expansion was uniform. In \fig{chap6:fig-cubeAnisotropic}(c) the increased $y$-direction stiffness of the first anisotropic model lead to a much reduced expansion in this direction (vertical in the image), and an accompanying increase in the orthogonal $z$-direction. For the second anisotropic model the direction defined by $y = z$ was stiffened, and \fig{chap6:fig-cubeAnisotropic}(d) shows the resulting reduced expansion along this axis and the increased orthogonal expansion. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic1.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic2.pdf}} \\
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic3.pdf}}
\hspace{1cm}
\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic4.pdf}} \\
\caption[Deformation patterns of transversely isotropic models compared with that of an isotropic model]{Deformation patterns of transversely isotropic models compared with that of an isotropic model: (a) the undeformed cube, (b) the deformed isotropic model, (c) the deformed anisotropic model with $ \mathbf{a}_0 = \mathbf{a}_0^y $, (d) the deformed anisotropic model with $\mathbf{a}_0^{yz}$. Colour maps indicate relative magnitude of lateral displacement  $((u_y^2 + u_z^2)^{1/2})$.}
\label{chap6:fig-cubeAnisotropic}
\end{figure}

@@

The end face reaction force history for each model is shown in \fig{chap6:fig-stressDecay}. A distinct decay curve resulting from stress relaxation, and similar to that commonly noted in biological tissues \citep{Fung93}, was observed. The increased stiffness afforded by the anisotropic models (albeit in an orthogonal direction to the loading) result in greater reaction forces than the isotropic model.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressDecay.pdf}
\end{center}
\caption[FE computed reaction forces on the end faces of the cube models over time.]{FE computed reaction forces on the end faces of the cube models over time. Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$. Note that in this figure both markers and solid lines correspond to FE solutions.}
\label{chap6:fig-stressDecay}
\end{figure}

@@

		\subsubsection*{Performance within SOFA}
The efficiency and the side-effects of porting the algorithm into the flexible SOFA framework need to be measured. Therefore the performance has been assessed by comparing the computational time of the algorithm running within and outside SOFA. NVIDIA provides a tool to check the GPU implementation by evaluating many variables during the execution like for instance timings, counts of inconsistent reads and writes or GPU occupancy. We used this tool to carry out two measures:
\begin{enumerate}
\item GPU time only estimates the GPU computational time.
\item CPU time allows the evaluation of the execution time with the additional overhead due to the framework.
\end{enumerate}
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=5.5cm]{chapter6/cube_and_curves1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[width=8cm]{chapter6/cube_and_curves2.pdf}}
\caption [Computational timings within SOFA] {(a) FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load. (b) Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.}
\label{chap6:fig-cube_and_curves}
\end{figure}
